Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-16 write-ups/PLDI/paper.tex
Spelling.

Notice that $k=0$ is always a valid choice for the number of
workgroups to be spawned by $\offerfork$, and is guarateed if $M$
is equal to the workgroup limit $N$.

@@

Our vision is that support for cooperative kernels will be integrated
in the runtimes of future GPU implementations of OpenCL and related
programming models, with driver and compiler support for our new
primitives.  To experiment with our ideas on current GPUs, we have
developed a prototype that mocks up the required runtime support via a
\emph{megakernel}, and exploits the occupancy-bound execution model
that these GPUs provide to ensure fair scheduling between workgroups.
We emphasise that an aim of cooperative kernels is to \emph{avoid}
depending on the occupancy-bound model.  Our protoytpe exploits this
model simply to allow us to experiment with current GPUs whose
proprietary drivers we cannot change.  We describe the megakernel approach (\mysec\ref{sec:megakernel})
and detail various aspects of the scheduler component of our implementation (\mysec\ref{sec:schedulerimpl}).

@@

For ease of implementation, our prototype supports
multitasking a single cooperative kernel with a single non-cooperative
kernel (though the non-cooperative kernel can be invoked many times).
We
assume all variables declared at the root lexical scope of a
cooperative kernel are implicitly $\transmit$-annotated; this allows our
implementation to piggy-back on top of Clang (see below) without
requiring front-end changes.  Finally, we require that $\offerkill$,
$\offerfork$ and $\resizingglobalbarrier$ are called from the entry
function of a cooperative kernel.  This allows us to use
$\keyword{goto}$ and $\keyword{return}$ to direct threads into and out
of the kernel.  With these restrictions we can experiment
with interesting irregular algorithms (see
\mysec\ref{sec:experiments}).  A non-mock implementation of
coopeartive kernels would not use the megakernel approach, so we do not deem
the engineering effort associated with lifting these restrictions to
be worthwhile.

@@

We consider barriers that use the master/slave model
(e.g.\ \cite{XF10}): one workgroup (the master) collects signals
from the other workgroups (the slaves) indicating that they have
arrived at the barrier and are waiting for a reply indicating that
they may leave the barrier. Once the master has recieved a signal from
all the slaves, it replies with a signal saying that the slaves may
leave.

@@

This barrier implementation is suboptimal becuase workgroups only
execute $\offerkill$ once per barrier call and, depending on
order of arrival, it is possible
that only one workgroup is killed per barrier call, preventing the scheduler from gathering workgroups quickly.

@@

\mytablong\ref{tab:applications} gives an overview of the 8 irregular
algorithms that we ported to cooperative kernels. Among them, 6 are
graph algorithms, based on the Pannotia~\cite{Pannotia} and
Lonestar~\cite{BNP12} GPU application suites, using global barriers.  We
indicate how many of the original number of barriers are changed to
resizing barriers, and how many variables need to be transmitted.  In
all cases all barriers were converted, except for bc which contains
barriers deeper in the call stack, a case not supported by our
prototype although these barriers could in principle be
converted. The remaining 2 algorithms are workstealing applications:
each required the addition of $\offerfork$ and $\offerkill$ at the start
of the main loop, and no variables needed to be transmitted.  For all
these examples the porting to cooperative kernels was similarly
straightforward to the porting illustrated in
\mysec\ref{sec:programmingguidelines} for our running examples. Most graph come with 2
different data sets as input, leading to 13 application/input pairs in
total.

@@

Among the GPUs available to us, four met our requirements: Intel HD 520,
HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux drivers
for OpenCL suffer from a known defect whereby long-running kernels lead
to defunct processes that the OS cannot
kill~\cite{amd-defunct-process}. We observed this on our AMD platform
with the latest driver, and simultaneously obtained alarming results on
this platform when using SVM atomics: modifying a kernel to include a
simple host/device handshake through SVM slowed down kernel execution by
an order of magnitude in some cases.  This behaviour, combined with the
defunt processes issue, lead us to suspect that the support we require
for cooperative kernels is not yet mature enough within AMD's drivers,
which we plan to discuss with engineers at AMD.

@@

We devised the following
procedure to determine non-cooperative OpenCL workloads that simulate
the computational intensity of various graphics rendering workloads.
%
We designed a synthetic kernel that occupies all workgrouops of a
GPU for a parameterised time period $t$, invoked in an infinite loop by a host application.  We then
searched for a maximum value for $t$ that allowed the
synthetic kernel to execute without having an observable impact on
graphics rendering.  Using the computed value, we
ran the host application for $X$ seconds, measuring the time $Y < X$
dedicated to GPU execution during this period and the number
of kernel launches $n$ that were issued.  We used $X \geq 10$ in all
experiments.  The values $(X-Y)/n$ and $X/n$ estimate the average time
spent using the GPU to render the display between kernel calls and the
period at which the OS requires the GPU for display rendering,
respectively.

@@

Our results are the following: light workloads needs to run 3ms every
70ms, medium ones 3ms every 40ms, and heavy ones 10ms every 40ms. For
the medium and heavy workloads, the 40ms period corresponds to graphics
rendering occuring 25 times per second, which coincides with the human
persistence of vision. The 3ms execution duration of both light and
medium configurations indicates that, as expected, GPU computation is
cheaper for basic display rendering compared with rendering that uses
complex shaders. We use these results to guide our evaluation of the
responsiveness of cooperative scheduling in
\mysec\ref{sec:responsiveness}.

@@

Our experiments span 13 pairs of cooperative kernels and inputs, 3
classes of non-cooperative kernel workloads, 4 quantities of workgroups
claimed for the non-cooperative kernel and 2 variations of resizing
barriers for graph algorithms, leading to 288 configurations.  We run
each configuration 10 times on each GPU, in order to report averaged
performance numbers. For each run, we record the execution time of the
cooperative kernel. For each scheduling of the non-cooperative kernel
during the run, we also record the \emph{gather time} needed by the
scheduler to collect enough workgroups to launch the non-cooperative
kernel, and the non-cooperative kernel execution time. We sanity-check
whether all kernels perform their computations corretly by comparing
their computed result with expected reference results.

@@

\paragraph{Performance} 
\myfiglong\ref{fig:performance} reports, for the Iris GPU, the overhead brought by the scheduling of
non-cooperative kernels over the cooperative kernel execution time.  This is the slowdown associated with running the cooperative kernel in the presence of multitasking, vs.\ running the cooperative kenrel in isolation (geometric mean over all applications and inputs).  We also show the period at which non-cooperative kernels can be scheduled (aritmetic mean over all applications
and inputs).
The results for the other GPUs are similar.
 We show
results for the three workloads listed in
\mysec~\ref{sec:sizingnoncoop}. The two gray horizontal lines in the
period graph correspond to the period goals of the workloads: the higher
(resp. lower) line corresponds to a period of 70ms (resp. 40ms) for the
light (resp. medium and heavy) workload.

@@

Still, the heavy
workloads meet their period (by allocating all-but-one workgroup for
non-cooperative kernels), incurring a slow down of less than 1.4$\times$ on average. Light
and medium workloads meet their period with only a negligable overhead
of the cooperative kernel (less than a 1.03$\times$ slowdown on average).

Overall we are very encouraged by these experimental findings, especially since they provide a lower bound on potential performance of our coperative kernels model. Implementing the model natively, with device-specific runtime support, could only improve performance and responsiveness compared with that of our megakernel-based prototype.

@@

Cooperative multitasking was offered in older operating systems
(e.g. Windows before the 95 edition) and is still used by some
operating systems, such as RISC
OS~\cite{risc-os-multitasking}.
Many programming languages offer cooperative multitasking via
variations of coroutines, e.g. coroutine objects in
Python
%\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
and Lua,
%\footnote{\url{https://www.lua.org/pil/9.1.html}}
and Ruby's
\emph{fiber}.
%\footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}

@@

In \mysec\ref{sec:semantics} we presented the semantics of cooperative
kernels relatively informally, using English, to provide the intuition
behind our programmming model.  We now back this up with a more formal
presentation as an operational semantics for an abstract GPU
programming model.

@@

GPUs are known to have relaxed memory models~\cite{ABDGKPSW-2015}.  To abstractly
account for this, we assume that the shared state component $\sigma$
is not simplyl a mapping from locations to values, but instead
captures all the intricacies of GPU memory that can lead to this
relaxed behaviour.  We also assume a function $\sync$ which, given a
kernel state $\kappa$, returns a set of kernel states.  The idea is
that each $\kappa' \in \sync(\kappa)$ is a possible kernel state that
can be reached from $\kappa$ by stalling until all stores to memory
and loads from memory to thread-local storage have completed.  All we
require is that $\sync$ does not modify the number of active
workgroups nor the component of a thread state that determines which
statements remain to be executed.

@@

Rule $\textsc{Kill-No-Op}$ reflects the fact that when a workgroup
reaches $\offerkill$, this can always be treated as a no-op.  Whether
a scheduler implementation accepts $\offerkill$ calls or not depends
on competing workloads and how the scheduler has been designed to meet
the non-functinoal requirements discussed in
\mysec\ref{sec:nonfunctional}, but in general the programmer should
always be prepared for the possiblity that a workgroup survives after
calling $\offerkill$.

@@

Recall that $\offerkill$ is a workgroup-level function: the same
syntactic $\offerfork$ call must be reached by all threads in a
workgroup.  This is captured in our rules by requiring in both
$\textsc{Kill-No-Op}$ and $\textsc{Kill}$ that every thread is ready
to execute $\offerkill$ followed by an identical statement $\myss$.
Neither rule applies until all threads in a workgroup reach
$\offerfork$, and the workgroup gets stuck if multiple threads in a
workgroiup reach $\offerfork$ with different subsequent statements.

@@

The $\textsc{Barrier}$ rule requires that \emph{every} thread
executing the kernel reaches $\globalbarrier()$ an identical following
statement.  This reflects that fact that $\globalbarrier()$ is a
kernel-level function.  Each thread then skips over the
$\globalbarrier()$ call, and the $\sync$ function is applied to yield
a set of kernel states that can arise due to memory synchronization
taking place.  An arbitary member of this set is selected as the next
kernel state.

@@

A disadvantage of this decision is that it may provide sub-optimal
responsiveness from the point of view of the scheduler.  Suppose the
scheduler requires an additional compute unit, but the active thread
with the larget id is processing some computationally intensive work
and will take a while to reach $\offerkill$.  Our semantics means that
the scheuduler cannot take advantage of the fact that another active
workgroup may invoke $\offerkill$ sooner.

@@

Our semantics dictate that the workgroup with id 0 will not be killed
if it invokes $\offerkill$.  This avoids the possibility of the
cooperative kernel terminating early due to the programmer
inadvertantly allowing all workgroups to be killed, and the decision
to keep workgroup 0 alive fits well with our choice to kill workgroups
in descending order of id.

@@

We opted for transmitting only selected varibles based on the
observation that many of a thread's private variables are dead at the
point of issuing $\offerfork$ or $\resizingglobalbarrier$, thus it
would be wasteful to transmit them.  A live variable analysis could
instead be employed to over-approximate the variables that might be
accessed by newly arriving workgroups, so that these are automatically
transmitted.

>>>>

Notice that $k=0$ is always a valid choice for the number of
workgroups to be spawned by $\offerfork$, and is guaranteed if $M$
is equal to the workgroup limit $N$.

@@

Our vision is that support for cooperative kernels will be integrated
in the runtimes of future GPU implementations of OpenCL and related
programming models, with driver and compiler support for our new
primitives.  To experiment with our ideas on current GPUs, we have
developed a prototype that mocks up the required runtime support via a
\emph{megakernel}, and exploits the occupancy-bound execution model
that these GPUs provide to ensure fair scheduling between workgroups.
We emphasise that an aim of cooperative kernels is to \emph{avoid}
depending on the occupancy-bound model.  Our prototype exploits this
model simply to allow us to experiment with current GPUs whose
proprietary drivers we cannot change.  We describe the megakernel approach (\mysec\ref{sec:megakernel})
and detail various aspects of the scheduler component of our implementation (\mysec\ref{sec:schedulerimpl}).

@@

For ease of implementation, our prototype supports
multitasking a single cooperative kernel with a single non-cooperative
kernel (though the non-cooperative kernel can be invoked many times).
We
assume all variables declared at the root lexical scope of a
cooperative kernel are implicitly $\transmit$-annotated; this allows our
implementation to piggy-back on top of Clang (see below) without
requiring front-end changes.  Finally, we require that $\offerkill$,
$\offerfork$ and $\resizingglobalbarrier$ are called from the entry
function of a cooperative kernel.  This allows us to use
$\keyword{goto}$ and $\keyword{return}$ to direct threads into and out
of the kernel.  With these restrictions we can experiment
with interesting irregular algorithms (see
\mysec\ref{sec:experiments}).  A non-mock implementation of
cooperative kernels would not use the megakernel approach, so we do not deem
the engineering effort associated with lifting these restrictions to
be worthwhile.

@@

We consider barriers that use the master/slave model
(e.g.\ \cite{XF10}): one workgroup (the master) collects signals
from the other workgroups (the slaves) indicating that they have
arrived at the barrier and are waiting for a reply indicating that
they may leave the barrier. Once the master has received a signal from
all the slaves, it replies with a signal saying that the slaves may
leave.

@@

This barrier implementation is sub-optimal because workgroups only
execute $\offerkill$ once per barrier call and, depending on
order of arrival, it is possible
that only one workgroup is killed per barrier call, preventing the scheduler from gathering workgroups quickly.

@@

\mytablong\ref{tab:applications} gives an overview of the 8 irregular
algorithms that we ported to cooperative kernels. Among them, 6 are
graph algorithms, based on the Pannotia~\cite{Pannotia} and
Lonestar~\cite{BNP12} GPU application suites, using global barriers.  We
indicate how many of the original number of barriers are changed to
resizing barriers, and how many variables need to be transmitted.  In
all cases all barriers were converted, except for bc which contains
barriers deeper in the call stack, a case not supported by our
prototype although these barriers could in principle be
converted. The remaining 2 algorithms are work stealing applications:
each required the addition of $\offerfork$ and $\offerkill$ at the start
of the main loop, and no variables needed to be transmitted.  For all
these examples the porting to cooperative kernels was similarly
straightforward to the porting illustrated in
\mysec\ref{sec:programmingguidelines} for our running examples. Most graph come with 2
different data sets as input, leading to 13 application/input pairs in
total.

@@

Among the GPUs available to us, four met our requirements: Intel HD 520,
HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux drivers
for OpenCL suffer from a known defect whereby long-running kernels lead
to defunct processes that the OS cannot
kill~\cite{amd-defunct-process}. We observed this on our AMD platform
with the latest driver, and simultaneously obtained alarming results on
this platform when using SVM atomics: modifying a kernel to include a
simple host/device handshake through SVM slowed down kernel execution by
an order of magnitude in some cases.  This behaviour, combined with the
defunct processes issue, lead us to suspect that the support we require
for cooperative kernels is not yet mature enough within AMD's drivers,
which we plan to discuss with engineers at AMD.

@@

We devised the following
procedure to determine non-cooperative OpenCL workloads that simulate
the computational intensity of various graphics rendering workloads.
%
We designed a synthetic kernel that occupies all workgroups of a
GPU for a parameterised time period $t$, invoked in an infinite loop by a host application.  We then
searched for a maximum value for $t$ that allowed the
synthetic kernel to execute without having an observable impact on
graphics rendering.  Using the computed value, we
ran the host application for $X$ seconds, measuring the time $Y < X$
dedicated to GPU execution during this period and the number
of kernel launches $n$ that were issued.  We used $X \geq 10$ in all
experiments.  The values $(X-Y)/n$ and $X/n$ estimate the average time
spent using the GPU to render the display between kernel calls and the
period at which the OS requires the GPU for display rendering,
respectively.

@@

Our results are the following: light workloads needs to run 3ms every
70ms, medium ones 3ms every 40ms, and heavy ones 10ms every 40ms. For
the medium and heavy workloads, the 40ms period corresponds to graphics
rendering occurring 25 times per second, which coincides with the human
persistence of vision. The 3ms execution duration of both light and
medium configurations indicates that, as expected, GPU computation is
cheaper for basic display rendering compared with rendering that uses
complex shaders. We use these results to guide our evaluation of the
responsiveness of cooperative scheduling in
\mysec\ref{sec:responsiveness}.

@@

Our experiments span 13 pairs of cooperative kernels and inputs, 3
classes of non-cooperative kernel workloads, 4 quantities of workgroups
claimed for the non-cooperative kernel and 2 variations of resizing
barriers for graph algorithms, leading to 288 configurations.  We run
each configuration 10 times on each GPU, in order to report averaged
performance numbers. For each run, we record the execution time of the
cooperative kernel. For each scheduling of the non-cooperative kernel
during the run, we also record the \emph{gather time} needed by the
scheduler to collect enough workgroups to launch the non-cooperative
kernel, and the non-cooperative kernel execution time. We sanity-check
whether all kernels perform their computations correctly by comparing
their computed result with expected reference results.

@@

\paragraph{Performance} 
\myfiglong\ref{fig:performance} reports, for the Iris GPU, the overhead brought by the scheduling of
non-cooperative kernels over the cooperative kernel execution time.  This is the slowdown associated with running the cooperative kernel in the presence of multitasking, vs.\ running the cooperative kernel in isolation (geometric mean over all applications and inputs).  We also show the period at which non-cooperative kernels can be scheduled (arithmetic mean over all applications
and inputs).
The results for the other GPUs are similar.
 We show
results for the three workloads listed in
\mysec~\ref{sec:sizingnoncoop}. The two gray horizontal lines in the
period graph correspond to the period goals of the workloads: the higher
(resp. lower) line corresponds to a period of 70ms (resp. 40ms) for the
light (resp. medium and heavy) workload.

@@

Still, the heavy
workloads meet their period (by allocating all-but-one workgroup for
non-cooperative kernels), incurring a slow down of less than 1.4$\times$ on average. Light
and medium workloads meet their period with only a negligible overhead
of the cooperative kernel (less than a 1.03$\times$ slowdown on average).

Overall we are very encouraged by these experimental findings, especially since they provide a lower bound on potential performance of our cooperative kernels model. Implementing the model natively, with device-specific runtime support, could only improve performance and responsiveness compared with that of our megakernel-based prototype.

@@

Cooperative multitasking was offered in older operating systems
(e.g. Windows before the 95 edition) and is still used by some
operating systems, such as RISC
OS~\cite{risc-os-multitasking}.
Many programming languages offer cooperative multitasking via
variations of co-routines, e.g. co-routine objects in
Python
%\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
and Lua,
%\footnote{\url{https://www.lua.org/pil/9.1.html}}
and Ruby's
\emph{fiber}.
%\footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}

@@

In \mysec\ref{sec:semantics} we presented the semantics of cooperative
kernels relatively informally, using English, to provide the intuition
behind our programming model.  We now back this up with a more formal
presentation as an operational semantics for an abstract GPU
programming model.

@@

GPUs are known to have relaxed memory models~\cite{ABDGKPSW-2015}.  To abstractly
account for this, we assume that the shared state component $\sigma$
is not simply a mapping from locations to values, but instead
captures all the intricacies of GPU memory that can lead to this
relaxed behaviour.  We also assume a function $\sync$ which, given a
kernel state $\kappa$, returns a set of kernel states.  The idea is
that each $\kappa' \in \sync(\kappa)$ is a possible kernel state that
can be reached from $\kappa$ by stalling until all stores to memory
and loads from memory to thread-local storage have completed.  All we
require is that $\sync$ does not modify the number of active
workgroups nor the component of a thread state that determines which
statements remain to be executed.

@@

Rule $\textsc{Kill-No-Op}$ reflects the fact that when a workgroup
reaches $\offerkill$, this can always be treated as a no-op.  Whether
a scheduler implementation accepts $\offerkill$ calls or not depends
on competing workloads and how the scheduler has been designed to meet
the non-functional requirements discussed in
\mysec\ref{sec:nonfunctional}, but in general the programmer should
always be prepared for the possibility that a workgroup survives after
calling $\offerkill$.

@@

Recall that $\offerkill$ is a workgroup-level function: the same
syntactic $\offerfork$ call must be reached by all threads in a
workgroup.  This is captured in our rules by requiring in both
$\textsc{Kill-No-Op}$ and $\textsc{Kill}$ that every thread is ready
to execute $\offerkill$ followed by an identical statement $\myss$.
Neither rule applies until all threads in a workgroup reach
$\offerfork$, and the workgroup gets stuck if multiple threads in a
workgroup reach $\offerfork$ with different subsequent statements.

@@

The $\textsc{Barrier}$ rule requires that \emph{every} thread
executing the kernel reaches $\globalbarrier()$ an identical following
statement.  This reflects that fact that $\globalbarrier()$ is a
kernel-level function.  Each thread then skips over the
$\globalbarrier()$ call, and the $\sync$ function is applied to yield
a set of kernel states that can arise due to memory synchronization
taking place.  An arbitrary member of this set is selected as the next
kernel state.

@@

A disadvantage of this decision is that it may provide sub-optimal
responsiveness from the point of view of the scheduler.  Suppose the
scheduler requires an additional compute unit, but the active thread
with the largest id is processing some computationally intensive work
and will take a while to reach $\offerkill$.  Our semantics means that
the scheduler cannot take advantage of the fact that another active
workgroup may invoke $\offerkill$ sooner.

@@

Our semantics dictate that the workgroup with id 0 will not be killed
if it invokes $\offerkill$.  This avoids the possibility of the
cooperative kernel terminating early due to the programmer
inadvertently allowing all workgroups to be killed, and the decision
to keep workgroup 0 alive fits well with our choice to kill workgroups
in descending order of id.

@@

We opted for transmitting only selected variables based on the
observation that many of a thread's private variables are dead at the
point of issuing $\offerfork$ or $\resizingglobalbarrier$, thus it
would be wasteful to transmit them.  A live variable analysis could
instead be employed to over-approximate the variables that might be
accessed by newly arriving workgroups, so that these are automatically
transmitted.
